---
title: "Problem 2"
author: "Maksym Stetskov, Mykyta Huskov"
date: "`r Sys.Date()`"
output: pdf_document
---

### PROBLEM 2 ---------------------------------------------------------------

## PART 2: Exploratory Data Analysis ----------------

**Imports**

```{r 2 setup, message= FALSE}
library(here)
library(tidyverse)
library(stargazer)
library(ggplot2)
library(zoo)
library(readxl)
library(stringr)

library(lmtest)
library(sandwich)
library(tseries)
library(knitr)
library(tinytex)
```

**Load the main dataset**

```{r 2 data, echo = FALSE}
data_houses <- read_excel(here("hw1", "data", "Ecox2 HW1 data New.xlsx"))
```

**Capture original column names.**
**Drop the time variable.**

```{r 2 capturing_names, include = FALSE}
true_names <- colnames(data_houses)
true_names <- true_names[-1]
```

**Rename columns for convinience**

```{r 2 renaming, include = FALSE}
colnames(data_houses) <- c("year_quarter", "const_overall", "constr_fam", "constr_apart", "renov_fam", "renov_apart", "constr_nondwell", "house_price_idx_2010", "pop", "cpi_yr_per", "nom_gdp")
```

### 2.2 Adding new variables from CNB and Eurostat data ----

**Import CNB and Eurostat sources.**

```{r 2 data_import,  echo = TRUE, message = FALSE}
cnb_loans_interest <- read_excel(here("hw1", "data", "cnb_loans_interest.xlsx"), col_names = FALSE, skip = 3)
cnb_loans_volume <- read_excel(here("hw1", "data", "cnb_loans_volume.xlsx"), col_names = FALSE, skip = 3)
eurostat_prod_prices <- read_excel(here("hw1", "data", "eurostat_prod_prices.xlsx"), col_names = FALSE, sheet = 2, skip = 10)
eurostat_employed <- read_excel(here("hw1", "data", "eurostat_employed.xlsx"), col_names = FALSE, sheet = 2, skip = 9)
```

**Normalize quarter formatting**

```{r 2 normalization_qrt}
data_houses$year_quarter <- as.yearqtr(gsub(" - ", " ", data_houses$year_quarter))
```

**Extract relevant columns from CNB data**

```{r 2 cnb_data}
cnb_loans_interest <- cnb_loans_interest[2:41, c(1, 2)] %>%
  rename_with(~ c("yr_qr", "house_loan_interest")) %>%
  arrange(yr_qr)
cnb_loans_volume <- cnb_loans_volume[2:41, c(1, 2)] %>%
  rename_with(~ c("yr_qr", "house_loan_volume")) %>%
  arrange(yr_qr)
```

**Attach CNB series to houses**

```{r 2 adding_cnb_to_main_df}
data_houses$house_loan_interest <- parse_double(cnb_loans_interest$house_loan_interest)
data_houses$house_loan_volume <- parse_double(cnb_loans_volume$house_loan_volume)
```

**Define Eurostat import helper**

```{r 2 eurostat_import}
eurostat_import <- function(eurostat_df, col_name = "value") {
  # Pull time and value rows
  time <- as.vector(t(eurostat_df[1, 2:ncol(eurostat_df)]))
  vals <- as.vector(t(eurostat_df[3, 2:ncol(eurostat_df)]))
  
  # Build output df
  out <- data.frame(yr_qr = time, value = vals) %>%
    dplyr::mutate(
      yr_qr = gsub("-", " ", yr_qr),
      yr_qr = as.yearqtr(yr_qr, format = "%Y Q%q"),
      value = na_if(value, ":"),
      value = parse_double(value)
    ) %>%
    tidyr::drop_na() %>%
    dplyr::filter(
      yr_qr >= as.yearqtr("2015 Q1", format = "%Y Q%q"),
      yr_qr <= as.yearqtr("2024 Q4", format = "%Y Q%q")
    ) %>%
    arrange(yr_qr)
  
  # Rename value to target name
  colnames(out)[2] <- col_name
  return(out)
}
```

**Import Eurostat employed and producer prices series**

```{r 2 importing_eurostat}
eurostat_employed <- eurostat_import(eurostat_employed, col_name = "employed_1000")
eurostat_prod_prices <- eurostat_import(eurostat_prod_prices, col_name = "prod_price_idx_2021")
```

**Append Eurostat series to houses.**

```{r 2 appending eurostat}
data_houses$employed_1000 <- eurostat_employed$employed_1000
data_houses$prod_price_idx_2021 <- eurostat_prod_prices$prod_price_idx_2021
```

**Adding invasion of Ukraine and quarter dummies**

```{r invasion_dummies}
data_houses <- data_houses %>%
  mutate(
    ones = 1L,
    quart = str_sub(as.character(year_quarter), -2),
  ) %>%
  pivot_wider(
    names_from = quart,
    values_from = ones,
    values_fill = 0,
    values_fn = sum
  )

data_houses <- data_houses %>%
  mutate(ukr_invasion = as.integer(year_quarter >= as.yearqtr("2022 Q1")))
```

**Cleaning intermediary objects**

```{r}
keep <- c("data_houses", "true_names")
rm(list = setdiff(ls(envir = .GlobalEnv), keep), envir = .GlobalEnv)
```

### 2.2.3 Summary table of the data  ----

```{r}
summary_stats <- function(df_col) {
  stats <- c(
    min(df_col, na.rm = TRUE),
    quantile(df_col, 0.25, na.rm = TRUE),
    median(df_col, na.rm = TRUE),
    mean(df_col, na.rm = TRUE),
    quantile(df_col, 0.75, na.rm = TRUE),
    max(df_col, na.rm = TRUE),
    sum(is.na(df_col))
  )
  return(stats)
}
```

```{r}
print(data_houses[10], n=40)
```

**Initialize summary table header.**

```{r}
summary_df <- data.frame(
  "Summary Statistics" = c(
    "Min", "First Quartile", "Median", "Mean", "Third Quartile", "Max",
    "Number of NAs"
  )
)
```

**Extend display names list.**

```{r}
true_names <- c(true_names, "Loans on house purchase (millions of CZK, current prices)", "Interest on house purchase loans", "Employed persons (thousands)", "Production price index (2021=100)")
```

**Pair display names with sources.**

```{r}
paired_names <- Map(list, true_names, colnames(data_houses)[-1])
```

**Fill summary table columns.**

```{r}
for (cols in paired_names) {
  new_name <- cols[[1]]
  old_name <- cols[[2]]
  
  df_col <- data_houses[[old_name]]
  
  if (is.numeric(df_col)) {
    summary_df[[old_name]] <- summary_stats(df_col)
  }
}
```

### 2.2.4 Visualizations of the most important variables  ----

**Histogram**

```{r}
# (No histogram code provided in the original block.)
```

**Time series plot with non-linear trend**

```{r}
data_houses %>%
  pivot_longer(
    cols = c(house_price_idx_2010, cpi_yr_per, house_loan_interest, const_overall, house_loan_volume, prod_price_idx_2021),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = year_quarter, y = value, group = variable)) +
  geom_line(aes(color = "Observed")) +
  geom_smooth(aes(color = "Linear trend"), method = "lm", se = FALSE) +
  facet_wrap(~variable, scales = "free_y") +
  scale_color_manual(values = c("Observed" = "black", "Linear trend" = "blue")) +
  guides(color = guide_legend(title = NULL)) +
  theme_bw() +
  labs(
    title = "Time series of the most important variables"
  )
```

## 2.2.5 Stationarity and trending variables ----

```{r}
data_houses %>%
  pivot_longer(
    cols = -c(year_quarter, Q1, Q2, Q3, Q4, ukr_invasion),
    names_to = "variable", values_to = "value"
  ) %>%
  ggplot(aes(x = year_quarter, y = value, group = variable)) +
  geom_vline(
    aes(
      xintercept = as.yearqtr("2022 Q1"),
      color = "Ukraine invasion"
    ),
    linewidth = 1.1
  ) +
  geom_line(aes(color = "Observed"), alpha = 0.9) +
  geom_smooth(aes(color = "Non-linear trend"), method = "loess", se = FALSE) +
  geom_point(aes(color = "Observed"), size = 0.8) +
  facet_wrap(~variable, scales = "free_y") +
  theme_classic() +
  scale_color_manual(
    values = c(
      "Observed" = "black",
      "Non-linear trend" = "blue",
      "Ukraine invasion" = "red"
    )
  ) +
  labs(
    title = "Stationarity and trending/seasonality analysis"
  )
```

## Part 3 ----

### 2.3.1 ----

```{r}
sm<-lm(data_houses$house_price_idx_2010~data_houses$nom_gdp)
summary(sm)

res_sm<-residuals(sm)
bp_sm<-lm(res_sm^2~data_houses$nom_gdp)
bptest(bp_sm)

bgtest(sm)

adf.test(data_houses$nom_gdp)
quarter <- substr(data_houses$year_quarter, 7, 7)
print(quarter)
gdp_sm_fdf <- lm(formula = data_houses$nom_gdp ~ as.factor(quarter), data = data_houses)
gdp_sm_fdf1 <- gdp_sm_fdf$residuals
gdp_sm_fdf2 <- c(NA,diff(gdp_sm_fdf1))
plot(gdp_sm_fdf2)

adf.test(data_houses$house_price_idx_2010)
```

### 2.3.2 ----

```{r}
gdp_detrend_ts <- ts(data_houses$nom_gdp, start = c(2015, 1), frequency = 4)  
gdp_detrend1 <- diff(gdp_detrend_ts, lag = 4)
print(gdp_detrend1)

hpi_detrend_ts <- ts(data_houses$house_price_idx_2010, start = c(2015, 1), frequency = 4)  
hpi_detrend <- diff(hpi_detrend_ts, lag = 4)
print(hpi_detrend)
```

### 2.3.3 ----

```{r}
sm_fdf<-lm(hpi_detrend_ts~gdp_detrend_ts)
summary(sm_fdf)
```

**correcting serial correlation**

```{r}
acf(residuals(sm_fdf))
dwtest(sm_fdf)

robust_se <- sqrt(diag(NeweyWest(sm_fdf, lag = 5, prewhite = FALSE)))
coeftest(sm_fdf, vcov = NeweyWest(sm_fdf, lag = 5, prewhite = FALSE))
```

## Part 4 ----

### 4.4.1 ----

```{r}
lnloanir<-log(data_houses$house_loan_interest)
data_houses$con_dwell<-data_houses$const_overall-data_houses$constr_nondwell

print(data_houses$con_dwell)
exp_in_model<-lm(data_houses$house_price_idx_2010~data_houses$nom_gdp+data_houses$employed_1000+data_houses$con_dwell+lnloanir+data_houses$house_loan_volume+data_houses$prod_price_idx_2021)
summary(exp_in_model)
```

### 4.4.2 ----

```{r}
acf(residuals(exp_in_model))
dwtest(exp_in_model)
bptest(exp_in_model)
```

**quarter correction**

```{r}
adf.test(data_houses$employed_1000)

emp_ts <- ts(data_houses$employed_1000, start = c(2015, 1), frequency = 4)  
emp_dt2 <- diff(emp_ts, lag = 4)
plot(emp_dt2)

adf.test(data_houses$con_dwell)

con_dwell_ts <- ts(data_houses$con_dwell, start = c(2015, 1), frequency = 4)  
con_dwell_dt2 <- diff(con_dwell_ts, lag = 4)
plot(con_dwell_dt2)

adf.test(lnloanir)

lnloanir_ts <- ts(lnloanir, start = c(2015, 1), frequency = 4)  
lnloanir_dt2 <- diff(lnloanir_ts, lag = 4)

plot(lnloanir_dt2)

adf.test(data_houses$house_loan_volume)

hlvl_ts <- ts(data_houses$house_loan_volume, start = c(2015, 1), frequency = 4)  
hs_dt2 <- diff(hlvl_ts, lag = 4)
plot(hs_dt2)
```

```{r}
det_exp_in_model<-lm(hpi_detrend~gdp_detrend1+emp_dt2+con_dwell_dt2+lnloanir_dt2+hs_dt2)
summary(det_exp_in_model)
```

**robust standard errors**

```{r}
acf(residuals(det_exp_in_model))

robust_se_1 <- sqrt(diag(NeweyWest(det_exp_in_model, lag = 2, prewhite = FALSE)))
coeftest(det_exp_in_model, vcov = NeweyWest(det_exp_in_model, lag = 2, prewhite = FALSE))
```

**third model**

```{r}
data_houses$ren_dwell<-data_houses$renov_fam-data_houses$renov_apart

rd_ts <- ts(data_houses$ren_dwell, start = c(2015, 1), frequency = 4)  
rend_detrend <- diff(rd_ts, lag = 4)
plot(rend_detrend)

ppi_ts <- ts(data_houses$prod_price_idx_2021, start = c(2015, 1), frequency = 4)  
ppi_detrend <- diff(ppi_ts, lag = 4)
plot(ppi_detrend)

micro_model<-lm(hpi_detrend~emp_dt2+con_dwell_dt2+lnloanir_dt2+rend_detrend+ppi_detrend)
summary(micro_model)
acf(residuals(micro_model))

acf(residuals(micro_model))
dwtest(micro_model)
bptest(micro_model)

robust_se_mm <- sqrt(diag(NeweyWest(micro_model, lag = 3, prewhite = FALSE)))
coeftest(micro_model, vcov = NeweyWest(micro_model, lag = 3, prewhite = FALSE))
```
